Efficient pedigree recording for fast population genetics simulation

### 2017-11-22 forwards_paper.tex
Updated introduction.

\begin{abstract}
    The computational burden of tracking individual genomes
    in a population genetics simulation can be substantial.
    In this note we describe how to both (a) dramatically reduce this burden and
    (b) efficiently record the entire history of the population.
    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
    on which neutral mutations can be quickly placed afterwards ---
    the \emph{tree sequence} introduced in the software package \msprime.
    We implement the method in two popular fowards-time simulation frameworks,
    and show it leads to order-of-magnitude speedups.
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g.\ $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent,
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20
% to assess the impact of European biased reference panels in genome wide association studies.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of interest in forwards-time, individual-based simulations.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals,
even though theory tells us that a population of size $N$
must be run for many multiples of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation
becomes quickly prohibitive.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.
Another way of representing this same information
is known as the \emph{ancestral recombination graph} ---
the {ARG} \citep{griffiths1997ancestral},
where nodes are common ancestor or recombination events ---
which has been the subject of substantial study
under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.

\jda{Is this the right place for this paragraph? should it maybe be integrated
	with graf beginning 'Modern computing' above?}
\plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
Forwards-time population genetics simulations are growing in importance,
as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity. 
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantiative traits under mutation-selection balance with
population sizes of $2 times 10^4$ diploids in stable populations and populations growing up to $\approx 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Due to tracking neutral mutations, the fastest simulation frameworks such as SLiM, SLiM2 \cite{haller2017flexible} and fwdpp \citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
implementations \cite{haller2017flexible}. In practice, current state-of-the-art simulation software may take on the
order of 
weeks to simulate models without selection\cite{fwdpp,Hernandez2015-wf}, and different simulation engines differ in how efficiently they
calculate fitnesses in models with selection \cite{fwdpp}.
These population and region sizes are still substantially short of whole genomes (hundreds to thousands of megabases)
for many biological population sizes of interest.
Simulating very large popluations and entire genomes will likely require parallelization, which \cite{lawrie2017accelerating} used to improve
performance for the "Poisson Random Field" family of models \cite{Sawyer1992-jw}, in which mutations are unlinked and there is no
effect of genetic background on fitness.  However, little, and perhaps even no, attention has been paid to
parallelization of simulations involving linkage.  Here, we show that recording genealogical information instead of
explicitly tracking neutral mutations results in order-of-magnitude performance improvements to simulations based on
\fwdpp{}.  The  idea of storing genealogical information to speed up simulations is not new.  AnA-FiTS implemented it
completely \citep{aberer2013rapid}, but this program did not implement the critical step of discarding irrelevant genealogical information.  A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.

In this paper we discuss storage methods for genealogies (and hence, genome sequence),
an algorithm for simplifying these,
and their use in forwards-time simulation.
Although we were motivated by a need for more efficient simulations,
these tools may prove more widely useful.

@@

>>>>

\begin{abstract}
    The computational burden of tracking individual genomes
    in a population genetics simulation can be substantial.
    In this paper we describe how to dramatically reduce this burden by
    efficiently recording the entire history of the population.
    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
    on which neutral mutations can be quickly placed afterwards ---
    the \emph{tree sequence} introduced in the software package \msprime.
    We implement the method in two popular fowards-time simulation frameworks,
    and show it leads to order-of-magnitude speedups.
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g.\ $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent,
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large scale forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantiative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\approx 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals,
even though theory tells us that a population of size $N$
must be run for many multiples of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation
becomes quickly prohibitive.
Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
implementations~\citep{haller2017flexible}.
In practice, current state-of-the-art simulation software may take on the order of
weeks to simulate models without selection~\citep{fwdpp,Hernandez2015-wf},
and different simulation engines differ in how efficiently they
calculate fitnesses in models with selection~\citep{fwdpp}.
These population and region sizes are still substantially short of whole genomes
(hundreds to thousands of megabases)
for many biological population sizes of interest.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.
In so doing we record the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time including the genetic outcomes of each ancestral
meiosis.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree
sequence}~\citep{kelleher2016efficient}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.
Another way of representing this same information
is known as the \emph{ancestral recombination graph} ---
the {ARG} \citep{griffiths1997ancestral},
where nodes are common ancestor or recombination events ---
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.

\jk{And finish with a more detailed descrition of what we do in the paper}
In this paper we discuss storage methods for genealogies (and hence, genome sequence),
an algorithm for simplifying these,
and their use in forwards-time simulation.
Although we were motivated by a need for more efficient simulations,
these tools may prove more widely useful.

@@

\jk{Moved this from the introduction: here seems like the right place to
discuss the importance of parallisation. We should rework this paragraph to
discuss say that a nice part of our approach is that we can do the
simplification in parallel. Also looking forward we need to parallise more
because the simulations need so much RAM}
Simulating very large popluations and entire genomes will likely require parallelization,
which \cite{lawrie2017accelerating} used to improve
performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
in which mutations are unlinked and there is no
effect of genetic background on fitness.  However, little, and perhaps even no, attention has been paid to
parallelization of simulations involving linkage.  Here, we show that recording genealogical information instead of
explicitly tracking neutral mutations results in order-of-magnitude performance improvements to simulations based on
\fwdpp{}.

\jk{Moved this here from intro also. Basically we just need to say that others
have done something similar. Ours is much better than Ana-Fits because ...}
The  idea of storing genealogical information to speed up simulations is not new.  AnA-FiTS implemented it
completely~\citep{aberer2013rapid}, but this program did not implement the critical step of
discarding irrelevant genealogical information.  A similar but more limited method for
discarding this information does appear in \citet{padhukasahasram2008exploring}.
